\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-14 paper/sp.tex
remaining to-dos all in-text

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figures 1 and 2 need to be combined to directly compare Parent and Training.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make the comparison with Cannon as a function of L1 strength and maybe look at correlation coefficients between parameter vectors.
% - Get model running on ASPCAP outputs and compare in Discussion.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email. And thank Ness.
% - Thank APOGEE telecon and add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Describe how we ran The Cannon; discussion of how we compare to The Cannon.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - write caption and paragraph about the new figure showing how much better we are than Gaia at large distance
% - need a snippet for Table 1.

@@

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
\begin{eqnarray}
\ssigma_m^2 &\leftarrow& \sparallax_m^2\,\hat{\theta}^T\cdot C_m\cdot\hat{\theta}
\end{eqnarray}
where $\ssigma_m$ is the uncertainty on the inferred (or predicted)
spectrophotometric parallax $\sparallax_m$,
the left-hand side is a scalar product,
and $C_m$ is the (presumed diagonal, actually) covariance matrix of the input
features.
That is, $C_m$ is a diagonal matrix with the uncertainty variances of the elements
of $x_m$ along the diagonal.

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44784 stars and is shown in \figurename~\ref{fig:parent}.
\begin{figure}
\includegraphics[width=\textwidth]{parent_sample.pdf}
\caption{Color--color diagrams for the 44784 Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.\label{fig:parent}}
\end{figure}

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
HOGG CITE STASSUN MAYBE AND OTHERS?
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the 28226 Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

Comparison of {\figurename s}~\ref{fig:parent} and \ref{fig:training} shows
that there is more color range in the Parent Sample than in the Training Set.
This challenges the representativeness assumption in \sectionname~\ref{sec:assumptions}.
The principal reason for the difference is that the \gaia\ quality cuts
exclude stars preferentially from crowded regions, which also tend to be the
most dust-obscured.
The hope of the model assumptions is that the Training Set will contain sufficient
dust variation that the model will naturally learn the dust corrections and extrapolate
acceptably.

@@

One strong assumption of this work is that our set of features is sufficient to
build a predictive model for distance.
Our argument for our included features is that they ought to be sufficient to
estimate a dust-independent apparent magnitude and the stellar parameters (especially
age and surface gravity) that are sensitive to luminosity.
In particular, since our model is linear in the logarithm of the parallax, it is
also linear in any kind of apparent magnitude at fixed distance and also linear
in distance modulus at fixed luminosity.
Our feature decisions were highly motivated by these kinds of considerations.
And the derived model parameters shown in \figurename~\ref{fig:spectral} do indicate that
the model is capturing some or all of the expected information in the spectra, at least.

@@

As we said above, our most restrictive assumption is the linearity assumption.
We required that the parallax prediction be constructed from a linear combination
of the input data or feature vector components.
This is absurd: Stars are not linear objects!
There are two questions to ask about the linearity.
One is: Why did it work? The other is: Could we have done better?
The answer to the first is that we have taken such a small part of stellar parameter
space with our surface gravity cut, that the linearized model of stellar
spectrophotometry is not a bad approximation.
Also, because our linear model included photometry in magnitudes (that is, logarithmic)
and was exponentiated to make the parallax prediction, we knew in advance that much
of the needed flexibility (for dust attenuation, for example) would be close to
linear.
That is, it was a combination of limited model scope and some cleverness in the
feature engineering and model structure.

The answer to the second question is that absolutely we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least non-worse---results.
The issues with going to a much more complex model are manifold, however:
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is close to linear; that is, if you give a model
a lot of freedom or capacity, it has to use a lot of the data to learn what part
of that capacity it really needs.
A more complex model might have had edge issues: Very flexible models don't extrapolate
outside the convex hull of their training data, and our training data was not
balanced in terms of parallax and dust attenuation.
And finally, the complex model would have been harder to visualize and interpret;
the linear model contains essentially only first derivatives in the data space, which are
relatively easy to understand.

>>>>

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email.
% - add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot and all references for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - write caption and paragraph about the new figure showing how much better we are than Gaia at large distance
% - need a snippet for Table 1.

@@

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
\begin{eqnarray}
\ssigma_m^2 &\leftarrow& \sparallax_m^2\,\hat{\theta}^T\cdot C_m\cdot\hat{\theta}
\label{eq:unc}
\end{eqnarray}
where $\ssigma_m$ is the uncertainty on the inferred (or predicted)
spectrophotometric parallax $\sparallax_m$,
the left-hand side is a scalar product,
and $C_m$ is the (presumed diagonal, actually) covariance matrix of the input
features.
That is, $C_m$ is a diagonal matrix with the uncertainty variances of the elements
of $x_m$ along the diagonal.

@@

We need complete feature-vector information for all stars.  For this
reason, we define the Parent Sample to be the set of all stars that
meet the \apogee\ and \gaia\ cuts and also
have the complete set of photometry: $G_n$, $\BP_n$, $\RP_n$, $J_n$,
$H_n$, $K_n$, $W_{1n}$, and $W_{2n}$ and also an \apogee\ spectrum.
In addition, we applied two light color cuts to remove stars with obviously
contaminated or outlying photometry:
\begin{eqnarray}
(J_n - K_n) &<& (+0.4\,\mathrm{mag}) + 0.45 * (\BP_n - \RP_n)
\\
(H_n - W_{2n}) &>& (-0.05\,\mathrm{mag})
\quad .
\end{eqnarray}
These cuts removed roughly 2~percent of the \apogee\ stars.
This Parent Sample contains 44784 stars and is shown in \figurename~\ref{fig:samples}.
\begin{figure}
\includegraphics[width=\textwidth]{./parent_sample.pdf}\\
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{\textsl{top:}
  Color--color diagrams for the 44784 Parent Sample stars.
  The points are colored by
  the \apogee\ pipeline estimates of surface gravity.
  The color cuts used to trim outliers are shown as grey lines.
  \textsl{bottom:}
  The same but for the 28226 Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:samples}}
\end{figure}

@@

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
HOGG CITE STASSUN MAYBE AND OTHERS?
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

Comparison of the top and bottom panels of {\figurename}~\ref{fig:samples} shows
that there is more color range in the Parent Sample than in the Training Set.
This challenges the representativeness assumption in \sectionname~\ref{sec:assumptions}.
The principal reason for the difference is that the \gaia\ quality cuts
exclude stars preferentially from crowded regions, which also tend to be the
most dust-obscured.
The hope of the model assumptions is that the Training Set will contain sufficient
dust variation that the model will naturally learn the dust corrections and extrapolate
acceptably.

@@

One strong assumption of this work is that our set of features is sufficient to
build a predictive model for distance.
Our argument for our included features is that they ought to be sufficient to
estimate a dust-independent apparent magnitude and the stellar parameters (especially
age and surface gravity) that are sensitive to luminosity.
In particular, since our model is linear in the logarithm of the parallax, it is
also linear in any kind of apparent magnitude at fixed distance and also linear
in distance modulus at fixed luminosity.
Our feature decisions were highly motivated by these kinds of considerations.

@@

As we said above, our most restrictive assumption is the linearity assumption.
We required that the parallax prediction be constructed from a linear combination
of the input data or feature vector components.
This is absurd: Stars are not linear objects!
There are two questions to ask about the linearity.
One is: Why did it work? The other is: Could we have done better?
The answer to the first is that we have taken such a small part of stellar parameter
space with our surface-gravity cut, that the linearized model of stellar
spectrophotometry is not a bad approximation.
Also, because our linear model included photometry in magnitudes (that is, logarithmic)
and was exponentiated to make the parallax prediction, we knew in advance that much
of the needed flexibility (for dust attenuation, for example) would be close to
linear.
That is, it was a combination of limited model scope and some cleverness in the
feature engineering and model structure.

The answer to the second question is that absolutely we could have done better!
Since models like Gaussian processes and deep networks subsume the linear model
(at least approximately), they could have delivered better---or at least non-worse---results.
The issues with going to a much more complex model are manifold, however:
We would have had many more decisions to make and many more hyper-parameters
to set.
We would have used some (or maybe much) of the information in the data to learn
the simple fact that the problem is close to linear; that is, if you give a model
a lot of freedom or capacity, it has to use a lot of the data to learn what part
of that capacity it really needs.
A more complex model might have had edge issues: Very flexible models don't extrapolate
outside the convex hull of their training data, and our training data was not
balanced in terms of parallax and dust attenuation.
In the  linear model it is also---unlike in more flexible models---trivial to propagate
uncertainties from the feature space to the prediction.
We used that, above, to propagate uncertainty in \equationname~(\ref{eq:unc}).
And finally---in principle---linear models are better for visualization and interpretation;
the linear model contains essentially only first derivatives in the data space, which are
relatively easy to understand.
